---
title: "Feature-informed Stage 10 omics workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feature-informed Stage 10 omics workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`forgeKI` can be run in two modes. The minimal mode designs HDR knock-in donors and guides without any cell-line omics resources. The feature-informed mode additionally builds a gene-specific Stage 10A-10E ranking from a local Stage 10 omics bundle. The omics bundle is external to the package because the underlying DepMap, CCLE, RRBS, and lab-derived HDR competency resources are large and should be versioned outside the source tree.

## 1. Minimal design run without omics

Use this mode to test guide enumeration, homology-arm construction, Type IIS handling, donor-module construction, virtual allele checks, and Stage 9 design scoring.

```{r, eval=FALSE}
library(forgeKI)

cfg <- forgeki_config(
  gene = "ACTB",
  project_dir = file.path(path.expand("~"), "forgeKI_runs", "ACTB_HDR"),
  donor = forgeki_donor_options(
    destination_vector_id = "pForge-Dest-HSVTK",
    fusion_module_id = "pForge-Fusion-HiBiT-p2A-EGFP",
    selectable_cassette_id = "pForge-Cassette-mRFP1-Hygro"
  ),
  guide = forgeki_guide_options(search_radius_bp = 80L, top_n = 10L),
  arms = forgeki_arm_options(lha_target_bp = 2000L, rha_target_bp = 2000L, min_arm_bp = 300L),
  golden_gate = forgeki_golden_gate_options(domestication_policy = "biology_first"),
  runtime = forgeki_runtime_options(save_rds = TRUE, write_progress = TRUE)
)

res <- run_forgeki_pipeline(
  cfg,
  job_root = file.path(cfg$project_dir, "jobs"),
  offtarget_mode = "exact_hg38",
  stage10_mode = "skip",
  top_n = 10L
)

report <- render_forgeki_report(res, output_dir = file.path(res$job$output_dir, "report"))
vendor <- export_forgeki_vendor_order_sheet(res, output_dir = file.path(res$job$output_dir, "vendor"))
```

This mode does not attempt to rank cell lines. It is the right entry point for validating the molecular design path.

## 2. Stage 10 resource-folder setup

A feature-informed Stage 10 run expects a clean input folder. A practical local layout is:

```text
forgeKI_omics_resources/
  inputs/
    20_HDR_CellLine_Ranking_Master.csv
    Model.csv
    depmap_rna_full.rds
    depmap_cn_full.rds
    depmap_crispr_full.rds
    depmap_mutations_full.rds
    OmicsFusionFiltered.csv
    CCLE_RRBS_TSS_1kb_20180614.txt
    CCLE_RRBS_TSS_CpG_clusters_20180614.txt
```

Create a resource checklist and quickstart artifacts with:

```{r, eval=FALSE}
input_dir <- file.path(path.expand("~"), "forgeKI_omics_resources", "inputs")
bundle_path <- file.path(path.expand("~"), "forgeKI_omics_resources", "forgeKI_stage10_omics_bundle.rds")

quick <- forgeki_stage10_resource_quickstart(
  input_dir = input_dir,
  output_dir = file.path(input_dir, "forgeKI_stage10_quickstart"),
  bundle_path = bundle_path,
  release_label = "forgeKI Stage 10 local omics bundle"
)

quick

check <- forgeki_check_stage10_omics_inputs(
  input_dir = input_dir,
  output_csv = file.path(input_dir, "forgeKI_stage10_quickstart", "stage10_omics_resource_check.csv")
)

check[, c("Resource_Key", "Expected_File", "Exists", "Requirement_Level", "Resource_Status")]
```

The quickstart writes a template CSV, an input-resource check CSV, a bundle README, and a compile script. Keep these files beside the input folder as the provenance record.

## 3. Compile and validate the RDS bundle

A single RDS bundle is the preferred runtime cache. The original raw files should still be retained for provenance.

```{r, eval=FALSE}
bundle <- forgeki_compile_stage10_omics_bundle(
  output_rds = bundle_path,
  depmap_root = input_dir,
  global_ranking_path = file.path(input_dir, "20_HDR_CellLine_Ranking_Master.csv"),
  cellline_metadata_path = file.path(input_dir, "Model.csv"),
  expression_path = file.path(input_dir, "depmap_rna_full.rds"),
  copy_number_path = file.path(input_dir, "depmap_cn_full.rds"),
  crispr_dependency_path = file.path(input_dir, "depmap_crispr_full.rds"),
  mutation_path = file.path(input_dir, "depmap_mutations_full.rds"),
  fusion_path = file.path(input_dir, "OmicsFusionFiltered.csv"),
  rrbs_tss_path = file.path(input_dir, "CCLE_RRBS_TSS_1kb_20180614.txt"),
  rrbs_cpg_path = file.path(input_dir, "CCLE_RRBS_TSS_CpG_clusters_20180614.txt"),
  release_label = "forgeKI Stage 10 local omics bundle",
  max_rows = Inf,
  compress = "gzip"
)

forgeki_validate_stage10_omics_bundle(bundle_path)
```

For runtime, `forgeKI` automatically materializes a per-gene slim bundle from the full bundle. The full bundle is the source archive; the gene-slim bundle is the fast working cache.

## 4. Full feature-informed run with Stage 10A-10E

Supply the bundle path through `forgeki_stage10_options()` and request the Stage 10 builder.

```{r, eval=FALSE}
cfg <- forgeki_config(
  gene = "ACTB",
  project_dir = file.path(path.expand("~"), "forgeKI_runs", "ACTB_HDR_stage10"),
  donor = forgeki_donor_options(
    destination_vector_id = "pForge-Dest-HSVTK",
    fusion_module_id = "pForge-Fusion-HiBiT-p2A-EGFP",
    selectable_cassette_id = "pForge-Cassette-mRFP1-Hygro"
  ),
  guide = forgeki_guide_options(search_radius_bp = 80L, top_n = 10L),
  arms = forgeki_arm_options(lha_target_bp = 2000L, rha_target_bp = 2000L, min_arm_bp = 300L),
  golden_gate = forgeki_golden_gate_options(domestication_policy = "biology_first"),
  stage10 = forgeki_stage10_options(
    top_n = 100L,
    omics_bundle_path = bundle_path,
    build_stage10_reference = TRUE,
    cellline_context_mode = "omics_builder",
    build_10a = TRUE,
    build_10b = TRUE,
    build_10c = TRUE,
    build_10d = TRUE,
    build_10e = TRUE
  ),
  runtime = forgeki_runtime_options(save_rds = TRUE, write_progress = TRUE)
)

res <- run_forgeki_pipeline(
  cfg,
  job_root = file.path(cfg$project_dir, "jobs"),
  offtarget_mode = "exact_hg38",
  stage10_mode = "auto",
  top_n = 10L
)

"stage10_reference_builder" %in% res$stages_completed
res$stages$stage10_reference_builder$stage10_final_summary

report <- render_forgeki_report(res, output_dir = file.path(res$job$output_dir, "report"))
```

The report audit directory should include:

```text
forgeKI_stage10_final_summary.csv
stage10_builder_feature_status.csv
stage10_builder_gene_feature_schema_audit.csv
stage10_builder_chromatin_schema_audit.csv
stage10_builder_practical_shortlist.csv
```

## 5. Interpreting Stage 10A-10E

Stage 10 is a decision-support layer. It prioritizes cell lines and designs; it does not guarantee editing success.

Stage 10A builds target-gene cell-line context from global HDR competency, model metadata, RNA expression, copy number, CRISPR dependency, mutation, and fusion resources. Stage 10B expands this context across candidate designs. Stage 10C carries allele-aware design context. Stage 10D overlays RRBS/chromatin proxy evidence when mappable. Stage 10E emits a final integrated ranking and a practical shortlist.

The most important summary fields are:

```text
Stage10_Context_Mode
Stage10A_QC_Status
Stage10E_QC_Status
N_Feature_Sources_Loaded
RRBS_Mapping_Status
N_RRBS_CellLines_Mapped
Top_CellLine_ID
Top_CellLine_Name
Top_Design_ID
Top_Final_Recommendation_Tier
Top_Final_Limiting_Factor_Summary
```

`Stage10_Context_Mode = feature_informed` means omics resources were used. `minimal` or global-only modes should be interpreted more conservatively.

## 6. What each feature contributes

The global HDR ranking provides a baseline cell-line competency prior. RNA expression flags whether the target gene is likely detectable in a given model. Copy number flags amplification or deletion context. CRISPR dependency flags whether perturbation of the target gene may affect fitness. Mutation and fusion evidence identify altered target-gene alleles that could affect reporter interpretation. RRBS is treated as a conservative chromatin/methylation proxy, not as a direct editing-efficiency measurement.

## 7. Troubleshooting

If Stage 10A reports `PASS_stage10a_context_constructed_minimal`, the run did not map gene-specific omics features. Check the feature-status and gene-feature schema-audit CSVs.

If RRBS reports `WARN_no_rrbs_locus_evidence_mapped`, the RRBS files were readable but not mappable to DepMap/cell-line IDs or to the target-gene locus. The Stage 10E ranking can still be produced, but chromatin evidence is unavailable.

If the full bundle is slow, use the normal pipeline path rather than manually loading and scanning the full bundle. The pipeline creates a per-gene slim runtime cache automatically.

If a required file is missing, rerun `forgeki_check_stage10_omics_inputs()` and inspect the `Resource_Status` and `Missing_Consequence` fields.
