---
title: "forgeKI v0.0.1 local pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{forgeKI v0.0.1 local pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`forgeKI` v0.0.1 provides a controlled local HDR reporter knock-in design pipeline. Each migrated stage can still be run independently, while `run_hdr_pipeline()` executes the gated stages in an isolated job directory and writes per-stage RDS/CSV outputs.

The current orchestrator runs Stage 1, Stage 2, Stage 4, Stage 5, Stage 6, Stage 7, Stage 8, Stage 3, and Stage 9. Stage 10 runs only when a fixed cell-line reference is supplied or explicitly required.

```{r, eval=FALSE}
library(forgeKI)

cfg <- hdr_config(
  gene = "ACTB",
  cassette_id = "toy_hibit",
  project_dir = "D:/Bioinformatics/HDR/forgeKI_Rpackage_test",
  guide = hdr_guide_options(search_radius_bp = 80L, top_n = 10L),
  arms = hdr_arm_options(lha_target_bp = 2000L, rha_target_bp = 2000L, min_arm_bp = 300L),
  runtime = hdr_runtime_options(save_rds = TRUE, write_progress = TRUE)
)

res <- run_hdr_pipeline(
  cfg,
  offtarget_mode = "none",
  stage10_mode = "skip",
  top_n = 10L
)

print(res)
res$stages_completed
res$stages$stage9_design_scoring$recommendation_summary
res$outputs
```

With `offtarget_mode = "none"`, Stage 9 is deliberately conservative and should not emit a primary production recommendation. Use this mode for pipeline validation and donor-construction checks, not final guide release.

Stage 10 consumes a fixed external cell-line reference. The package tests use toy references for portability; production use should point to the lab-maintained `20_HDR_CellLine_Ranking_Master.rds` or `.csv` file. The broader scoring bundle should be retained for provenance/audit.

```{r, eval=FALSE}
cell_ref <- load_hdr_cellline_reference(
  "D:/Bioinformatics/HDR/HDR_Stage2_CellLine_Ranking/20_HDR_CellLine_Ranking_Master.rds"
)

res10 <- run_hdr_pipeline(
  cfg,
  cellline_reference = cell_ref,
  offtarget_mode = "none",
  stage10_mode = "auto",
  top_n = 10L
)

res10$stages$stage10_cellline_context$cellline_context
```

The same reference can also be configured persistently:

```{r, eval=FALSE}
cfg <- hdr_config(
  gene = "ACTB",
  cassette_id = "toy_hibit",
  project_dir = "D:/Bioinformatics/HDR/forgeKI_Rpackage_test",
  stage10 = hdr_stage10_options(
    cellline_reference_path = "D:/Bioinformatics/HDR/HDR_Stage2_CellLine_Ranking/20_HDR_CellLine_Ranking_Master.rds",
    require_cellline_reference = TRUE
  )
)

res10 <- run_hdr_pipeline(cfg, offtarget_mode = "none", stage10_mode = "auto")
```

If no direct object or config path is supplied, the orchestrator checks the `FORGEKI_CELLLINE_REFERENCE` environment variable, with `HDRDESIGNR_CELLLINE_REFERENCE` retained as a legacy alias.

A typical job directory contains one subdirectory per stage under `outputs/stages`, a final `hdr_result.rds`, progress logs, and a manifest.
